---
title: "Class 1"
author: "Josephine Uwamariza"
date: "2025-10-16"
output: html_document
---
#Vector: The foundational R data structure
#Numeric
```{r}
x <- c(0.5, 0.6)
age<-c(20,35,32,29)
summary(age)
mean(age)
```
#Logical
```{r}
x <- c(TRUE, FALSE)
x <- c(T, F)
```
#Character
```{r}
x <- c("a", "b", "c" )
class<-c("M","F","F","M")
```
#Integer
```{r}
x <- 9:29
```
#Complex
```{r}
x <- c(1+0i, 2+4i)
```
#Matrix and operation within matrices
```{r}
x<- matrix(1:6, nrow = 2, ncol = 3)
x
```
```{r}
(y<-matrix(1:6, nrow = 3, ncol = 2))
```
```{r}
x%*%y
```
#List
```{r}
x <- list(1, "a", TRUE, 1 + 4i)
```
#Factor
```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
x
```
#Create a vector with NAs in it
```{r}
x <- c(1, 2, NA, 10, 3)
```
#Return a logical vector indicating which elements are NA
```{r}
is.na(x)

x <- c(1, 2, 4, "NA", 5)
bad <- is.na(x)
print(bad)
x[!bad]
```
#What if there are multiple R objects and you want to take the subset with no missing
#values in any of those objects?
```{r}
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x, y)
good
```
#Coercion
#If character is present, in a vector, R convert everything in the vector to character
#strings.
#If a vector only contains logical and numbers, R will convert the logical to numbers,
#Every true becomes a 1, and every FALSE becomes 0
```{r}
sum(c(TRUE,TRUE,FALSE,FALSE,FALSE))
```
#Create data
#Create data Using data frame Data frame is more general than matrix How???
#Because data frame can contain different modes of data (Numeric, character and so
#on) Similar to what you can see in SPSS, SAS,…

#Let’s create a data frame
```{r}
studentID<-c(1,2,3,4,5)
math_score<-c(12,17,10,9,NA)
gender<-c("M","F","M","M","F")
it_score<-c(13,18,11,10,19)
scoredata<-data.frame(studentID,gender,math_score,it_score)
scoredata
```

#Create data from keyboard
#Steps 1. Create data frame (or matrix) with variable names 2. Invoke the text editor
#in the data objected created at first step

```{r}
data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric
(0))
data_class2<-edit(data_class2)
```
#Data Importation
#R has some features that can allow to import data from different sources ( It can be
#text file, spreadsheet, or database)
#1. Data from excel
#2. Data statistical packages (SAS, SPSS, Stata)
#3. Data from Text files (ASCII, XML, Webscraping)
#4. Data from database management systems (SQL,MySQL, Oracle, Access)

#HW1: Import data from Statistical package and from Database management systems
```{r}
world_population <- read.csv("C:\\Users\\HP\\Desktop\\R Programming\\Mid Term Exam\\world_population.csv")
variable.names(world_population)

```
```{r}
head(world_population)
```

```{r}
tail(world_population)
```

```{r}
world_population[10:20,]
```

```{r}
colnames(world_population)
summary(world_population$Capital)
```
```{r}
length(world_population$Capital)
```
```{r}
world_population[,-1]
```

#In case you want to use data set built in R
```{r}
data() # list of datasets currently available
data("airquality")
variable.names(airquality)
```
```{r}
str(airquality)
```
#Exploration data analysis
#Examine numerical variable using common summary
```{r}
world_population
```

```{r}
summary(world_population)
```
```{r}
str(world_population)
```
#Summary Statistics
```{r}
summary(world_population)
```

```{r}
apply(world_population,2,mean)
```
```{r}
apply(world_population,2,sd)
```

```{r}
c(mean(world_population$X2020.Population),sd(world_population$X2020.Population))
```

```{r}
c(mean(world_population$X1990.Population),sd(world_population$X1990.Population))
```

```{r}
c(Mean=mean(world_population$X2020.Population),SD=sd(world_population$X2020.Population))
```

```{r}
c(Mean=mean(world_population$X1990.Population),SD=sd(world_population$X1990.Population))
```
#Variation and covariation using both numerical and graphs

#Relationship between a categorical and a continuous variable

```{r}
data("iris")
head(iris)
```
```{r}
mean(iris$Petal.Length)
```
```{r}
mean(iris$Petal.Length[iris$Species=="setosa"])
```
```{r}
mean(iris$Petal.Length[iris$Species=="versicolor"])
```
```{r}
mean(iris$Petal.Length[iris$Species=="virginica"])
```
## shortcut

```{r}
by(iris$Petal.Length,iris$Species,mean)
```

```{r}
by(iris$Petal.Length,iris$Species,sd)
```
```{r}
by(iris$Petal.Length,iris$Species,summary)
```
#Visualize the differences in continuous variables between categories using Box-andwhisker
#plot
```{r}
boxplot(iris$Sepal.Length~iris$Species,data=iris, main="Comparison")
```
#Relationship between two categorical variables
```{r}
ucba<-data.frame(UCBAdmissions)
head(ucba)
```
```{r}
cross<-xtabs(Freq~Gender+Admit,data=ucba)
(cross<-xtabs(Freq~Gender+Admit,data=ucba))
```
## Is there gender bias in UCB graduate admission process?

```{r}
prop.table(cross,2)
```
#Simpson’s paradox

#Phenomenon, where a trend that appears in combined groups of data disappears or
#reverses when broken down into groups.
```{r}
cross2<-xtabs(Freq~Gender+Admit,data=ucba[ucba$Dept=="A",])
prop.table(cross2,1)
```
#Bar Chart
```{r}
data <- read.table(text ="ProdA ProdB ProdC ProdD
1 110 50 60 70
2 120 50 80 65", header= TRUE)

barplot(as.matrix(data),beside=FALSE,col=c("brown","skyblue"))
```
#barplot(as.matrix(dat),beside=TRUE,col=c("gold3","red"))
```{r}
data <- read.table(text = "A B C D E F G
1 10 80 30 90 70 60 90
2 20 50 70 50 40 10 40
3 60 80 80 60 60 30 160
4 20 40 70 80 20 10 70", header = TRUE)
barplot(as.matrix(data))
```
#Scatter Plot
```{r}
ChickWeight
colnames(ChickWeight)
```
```{r}
plot(weight~Time,data=ChickWeight)

```
```{r}
plot(ChickWeight$Time,ChickWeight$weight)
```
```{r}
plot(ChickWeight[,1],ChickWeight[,2])
```
```{r}
cor(ChickWeight$Time,ChickWeight$weight)
```
#Linear Models
```{r}
plot(weight~Time,data=ChickWeight)
abline(lm(weight~Time,data=ChickWeight)$coefficient)
```
```{r}
data_lm<-lm(weight~Time,data=ChickWeight)
summary(data_lm)
```
```{r}
hist(ChickWeight$weight)
```
```{r}
hist(ChickWeight$weight,col = "skyblue")
```
#Linear with other line
```{r}
plot(weight~Time,data=ChickWeight)
abline(lm(weight~Time,data=ChickWeight)$coefficient)
lines(lowess(ChickWeight$Time,ChickWeight$weight),col="red")
```
```{r}
cor(ChickWeight$Time,ChickWeight$weight)
```
```{r}
model1<-lm(weight~Time,data=ChickWeight)
```
#Data management

```{r}
data_class<-read.table("C:\\Users\\HP\\Desktop\\R Programming\\Class data.txt")
```
##Create another Variable
```{r}
attach(data_class)
(BMI<-WEIGHT/(HEIGHT/100)^2)
```



